---
title: "Functions to work with text"
# subtitle: "Biostatistical Computing with R"
# author: "Mikhail Dozmorov"
# institute: "Virginia Commonwealth University"
# date: "`r Sys.Date()`"
# date: "08-23-2023"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["xaringan-themer.css", "xaringan-my.css"]
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false

---

```{r xaringan-themer, include = FALSE}
options(blogdown.knit.serve_site = FALSE)
library(xaringanthemer)
mono_light(
  base_color = "midnightblue",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "500", "500i"),
  code_font_google   = google_font("Droid Mono"),
  link_color = "#8B1A1A", #firebrick4, "deepskyblue1"
  text_font_size = "28px",
  code_font_size = "26px"
)
```

## Text format

- Text is the most cross-platform format to share data

- Typically, data is stored as field-delimited columns (think Excel). Delimiter may be tab character (".tsv" or ".txt" file extension), or comma (comma-separated values, ".csv")

- Disadvantage - can be large. Solution - compression (**gzip**ping), with tools to manipulate compressed files without uncompressing

- Text data often needs to be cleaned and formatted/restructured before analysis

- Extracting meaningful information from text requires pattern matching using regular expressions (regex)

---
## Examples of string manipulation

- **Base R Functions**: Functions like `grep()`, `gsub()`, `strsplit()`, and more

```{r}
text <- "The quick brown fox jumps over the lazy dog."
print(strsplit(text, " ")[[1]])
```

- **Stringr Package**: Part of the Tidyverse, offers a consistent set of functions designed for string manipulation (`str_replace()`, `str_split()`, etc.)

```{r message=FALSE, warning=FALSE}
library(stringr)
print(str_split(text, " ")[[1]])
```

---
## Basic String Operations

- `nchar()` - Counting Characters in a String
```{r}
text <- "Biostatistics"
print(nchar(text))  
```
- `toupper()`, `tolower()` - Changing Case
```{r}
print(toupper(text))
```

---
## Basic String Operations

- `substr()` - Extracting or Replacing Substrings
```{r}
text <- "Biostatistics is fascinating"
print(substr(text, 1, 4))
```

- `strsplit()` - Splitting Strings into Substrings by a delimiter. 
Returns a list containing split words

```{r}
print(strsplit(text, " ")[[1]]) # The content, split words
```

---
## Concatenation strings

- `paste()`: Concatenates strings with a separator
```{r}
text1 <- "Bio"
text2 <- "statistics"
print(paste(text1, text2, sep = "-"))
print(paste(c(text1, text2), sep = "-")) # Doesn't work for character vectors
print(paste(c(text1, text2), collapse = "-"))
```

---
## Concatenation strings

- `paste0()`: Concatenates strings without any separator
```{r}
text1 <- "Bio"
text2 <- "statistics"
print(paste0(text1, text2)) # Default separator is "", nothing
print(paste0(c(text1, text2))) # Still need to use collapse = " "
```

---
## Introduction to Regular Expressions (regex)

- Regular expressions (regex) are sequences of characters that form search patterns

- Used for string matching, searching, and replacing within text data

- Data cleaning, validation, extraction, and transformation

.small[ https://en.wikipedia.org/wiki/Regular_expression#POSIX_basic_and_extended ]

---
## Introduction to Regular Expressions (regex)

**Metacharacters**: .^$*+?{}[]\|()
- `.` - Matches any character except newline
  - `c.t` - Matches any string containing 'c', any character, then 't' (e.g., 'cat', 'cot')
- `^` - Matches the start of a string
  - `^A` - Matches any string that starts with 'A'
- `$` - Matches the end of a string
  - `dog$` - Matches any string that ends with 'dog'
- `*` - Matches 0 or more repetitions
- `+` - Matches 1 or more repetitions
- `?` - Matches 0 or 1 repetition
- `[]` - Matches any one of the characters inside the brackets
- `|` - Matches either the pattern before or the pattern after the |

---
## Using `grep()`, `grepl()` for Pattern Matching and Replacement

- `grep()` - Returns indices of the elements that match the pattern
- `grepl()` - Returns a logical vector indicating if a pattern was found
```{r}
text <- c("cat", "dog", "bird", "catfish")
print(grep("^cat", text))
print(grepl("cat$", text))
```

---
## Using `sub()`, and `gsub()` for Pattern Matching and Replacement

- `sub()` - Replaces the first instance of a pattern in a string
```{r}
text <- "The cat sat on the cat mat."
print(sub("cat", "dog", text))
```
- `gsub()` - Replaces all instances of a pattern in a string (global sub)
```{r}
print(gsub("cat", "dog", text))
```

---
## String Formatting

- `formatC()` - Creating Formatted Strings. Also, `sprintf()`
  - Syntax: `formatC(x, format, digits, flag, width, ...)`
```{r}
numbers <- c(123.456, 78.9, 0.1234)
# Fixed-point formatting
print(formatC(numbers, format = "f", digits = 2))
# Scientific notation
print(formatC(numbers, format = "e", digits = 2))
```

---
## String Formatting

- `formatC()` - Creating Formatted Strings. Also, `sprintf()`
  - Syntax: `formatC(x, format, digits, flag, width, ...)`
```{r}
numbers <- c(123.456, 78.9, 0.1234)
# Padding numbers with zeros
print(formatC(numbers, format = "d", flag = "0", width = 5))
# Left and right alignment, note string conversion
print(formatC(numbers, format = "d", flag = "-", width = 5))
```

---
## String Manipulation with stringr Package

Functions in `stringr` for enhanced string manipulation
- `str_sub()`, `str_replace()`, `str_remove()`, `str_to_upper()` and more

https://stringr.tidyverse.org/

